Extended Data

### 2022-03-22 doc/manuscript/Dow_manuscript_Nature.Rmd
reduced refs down to 50!
#125 #129

**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021<!--;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;-->@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
**However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum growth rates, or annual growth.**
**The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
**These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity enough to strengthen the long-term CO~2~ sink of temperate deciduous forests.**

@@

In recent decades, tree growth in Earth's forests has more than offset losses from deforestation and other disturbances, such that a net forest CO~2~ sink of ~`r round(5.8*12/44,1)` Gt C yr^-1^<!--net forest estimate in Harris et al. Table 2--> offsets ~`r round(5.8/32,1)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al. Table 2)--> of anthropogenic emissions[@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and climate change.
<!--@xu_changes_2021 provides an alternate estimate, which is about half of Harris. The main difference being that Xu includes savannas and woodlands, which have experienced a lot of loss from fires. Given focus on forests here, Harris seems more appropriate. -->
Of this important C sink, ~ `r round(3.6/7.6,2)*100`%<!-- temperate net / global net, (Harris et al. Table 1)--> occurs in temperate forests[@harris_global_2021],
<!-- alternate estimate: `r round(.31/.88,2)*100`% (temperate net / global net, (@xu_changes_2021. Table 2))--> 
with temperate deciduous forests sequestering >0.3 Gt C yr^-1^ [@pugh_role_2019]. <!-- Pugh et al. 2019 estimate that the temperate deciduous sink is >300 Tg C yr-1, with ~200 Tg C yr-1  attributable to demography (forest regrowth) and >100 Tg C yr-1  controlled by environmental drivers.--> 
The future behavior of this C sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020;@friedlingstein_global_2020;@fatichi_moving_2014].

@@

Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;<!--@delpierre_chilling_2019;-->@rossi_critical_2008] and increased annual growth in mesic climates[<!--@cook_predicting_1991;-->@babst_twentieth_2019;<!--@anderson-teixeira_joint_2022;--> @gao_earlier_2022]. 
In contrast, we have little evidence as to how stem growth and ANPP~woody~ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021<!--;@delgado_differences_2020-->].<!--Most of htese citations are important in other spots, but do we need all of them here?-->
However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], gross primary productivity (GPP)[@delpierre_wood_2016], or NEE[@delpierre_wood_2016]. 
Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zani_increased_2020;@zohner_how_2021].
Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest, but usually insignificant, positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014;@lu_no_2022].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

@@

<!--summary, setting expectation of extra C-->
Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`-`r tree_rings`). 
The observed advance in stem growth under warmer springs parallels advances observed for canopy foliage phenology (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and NEE[@keenan_net_2014;@finzi_carbon_2020]. 
However, inconsistent with the concept that an earlier start to growth would increase ANPP~woody~, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig. `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`).
Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014;-->@delpierre_wood_2016;@delpierre_temperate_2016].
This adds to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@fatichi_moving_2014], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, do not cause a corresponding increase in stem growth[@etzold_number_2022;@jiang_fate_2020]. <!--rocha_linking_2006 may work here if needed-->

>>>>

**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;<!--@stuble_plant_2021;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;-->@jeong_phenology_2011;@finzi_carbon_2020;<!--@stuble_plant_2021-->] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
**However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum growth rates, or annual growth.**
**The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
**These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity enough to strengthen the long-term CO~2~ sink of temperate deciduous forests.**

@@

In recent decades, tree growth in Earth's forests has more than offset losses from deforestation and other disturbances, such that a net forest CO~2~ sink of ~`r round(5.8*12/44,1)` Gt C yr^-1^<!--net forest estimate in Harris et al. Table 2--> offsets ~`r round(5.8/32,1)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al. Table 2)--> of anthropogenic emissions[@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and climate change.
<!--@xu_changes_2021 provides an alternate estimate, which is about half of Harris. The main difference being that Xu includes savannas and woodlands, which have experienced a lot of loss from fires. Given focus on forests here, Harris seems more appropriate. -->
Of this important C sink, ~ `r round(3.6/7.6,2)*100`%<!-- temperate net / global net, (Harris et al. Table 1)--> occurs in temperate forests[@harris_global_2021],
<!-- alternate estimate: `r round(.31/.88,2)*100`% (temperate net / global net, (@xu_changes_2021. Table 2))--> 
with temperate deciduous forests sequestering >0.3 Gt C yr^-1^ [@pugh_role_2019]. <!-- Pugh et al. 2019 estimate that the temperate deciduous sink is >300 Tg C yr-1, with ~200 Tg C yr-1  attributable to demography (forest regrowth) and >100 Tg C yr-1  controlled by environmental drivers.--> 
The future behavior of this C sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@ahlstrom_robustness_2012;@friedlingstein_global_2020;@fatichi_moving_2014].

@@

Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;<!--@delpierre_chilling_2019;-->@rossi_critical_2008] and increased annual growth in mesic climates[<!--@cook_predicting_1991;-->@babst_twentieth_2019;<!--@anderson-teixeira_joint_2022;--> @gao_earlier_2022]. 
In contrast, we have little evidence as to how stem growth and ANPP~woody~ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;<!--@stuble_plant_2021;@delgado_differences_2020-->].<!--Most of htese citations are important in other spots, but do we need all of them here?-->
However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], gross primary productivity (GPP)[@delpierre_wood_2016], or NEE[@delpierre_wood_2016]. 
Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zani_increased_2020;@zohner_how_2021].
Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest, but usually insignificant, positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014;@lu_no_2022].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

@@

<!--summary, setting expectation of extra C-->
Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`-`r tree_rings`). 
The observed advance in stem growth under warmer springs parallels advances observed for canopy foliage phenology (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020<!--;@stuble_plant_2021-->] and NEE[@keenan_net_2014;@finzi_carbon_2020]. 
However, inconsistent with the concept that an earlier start to growth would increase ANPP~woody~, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig. `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`).
Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014;-->@delpierre_wood_2016;@delpierre_temperate_2016].
This adds to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@fatichi_moving_2014], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, do not cause a corresponding increase in stem growth[@etzold_number_2022;@jiang_fate_2020]. <!--rocha_linking_2006 may work here if needed-->
